On Solving the Problem of Finding Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction Using Parallel Global Search Algorithm

### 2022-09-22 2022/PAVT Special Issue/template.tex
submitted version of the paper

%=================================================================
% Full title of the paper (Capitalized)
\Title{Using the ... Algorithm to Detect Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction}
% Дежурное название: Использование ...алгоритма для определения кинетических параметров процесса каталитической изомеризации пентан-гексановой фракции
% Еще один вариант: О решении задачи определения Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction с помощью Parallel Global Search Algorithm
% На английском: About Solving the Problem of Finding Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction Using Parallel Global Search Algorithm

% MDPI internal command: Title for citation in the left column
\TitleCitation{Title}

% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0000-0001-5273-2471} % Add \orcidA{} behind the author's name
\newcommand{\orcidauthorB}{0000-0002-8736-0652} % Add \orcidB{} behind the author's name
\newcommand{\orcidauthorC}{0000-0003-4219-4870}

% Authors, for the paper (add full first names)
\Author{Konstantin Barkalov $^{1}$\orcidA{}, Irek Gubaydullin $^{2,3}$, Ilya Lebedev $^{1}$\orcidB{}, Roza Faskhutdinova $^{2,3}$ and Leniza Enikeeva $^{3,4,}$*\orcidC{}}

@@

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{Konstantin Barkalov, Irek Gubaydullin, Ilya Lebedev, Roza Faskhutdinova and Leniza Enikeeva}

% MDPI internal command: Authors, for citation in the left column
\AuthorCitation{Barkalov, B.; Gubaydullin, I.; Lebedev, I.; Faskhutdinova, R.; Enikeeva, L.}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{A single paragraph of about 200 words maximum. For research articles, abstracts should give a pertinent overview of the work. We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations. The abstract should be an objective representation of the article, it must not contain results which are not presented and substantiated in the main text and should not exaggerate the main conclusions.}

% Keywords
\keyword{kinetic model; inverse problems of chemical kinetics; mathematical modeling; catalytic isomerisation of pantane-hexane fraction (List three to ten pertinent keywords specific to the article; yet reasonably common within the subject discipline.)}

@@

%Часть ННГУ - начало

@@

Numerical methods for solving such multiextremal problems (global optimization methods) differ significantly from local optimization methods since they require the study of the entire search domain (see, for example, \cite{PaulaviciusZilinskas2014,Sergeyev2017}). These methods can be divided into two classes: meta-heuristic and deterministic. Metaheuristic algorithms are usually based on simulations of processes occurring in the wild.
Examples of metaheuristic algorithms are simulated annealing, evolution and genetic algorithms, etc. (see, for example, \cite{Battiti2009,Eiben2015}).
These algorithms are included in many software libraries for solving global optimization problems (for example, MATLAB Global Optimization Toolbox, Scipy.Optimize, etc.).
Due to the operation speed and low requirements for computing resources (in particular, memory), metaheuristic algorithms have become widespread for solving practical problems. However, the solution to the problem found by such an algorithm is, in general, local and can be located far from the global optimum \cite{Kvasov2018}.

@@

This paper reflects the results of applying parallel methods of Lipschitz optimization developed in \cite{Barkalov2016,Strongin2018,Sysoyev2017}, to the solution of inverse problems of chemical kinetics.  
The main part of the paper is structured as follows. 
Section 2 provides a description of the mathematical model of the chemical reaction under study. Section 3 discusses the standard formulation of the Lipschitz global optimization problem and the algorithm for solving it. It also outlines the issues related to the technical aspects of the implementation of the parallel algorithm. The results of the numerical solution of the inverse problem of chemical kinetics using the developed parallel algorithm are discussed in Section 4. The last section contains the conclusion and plans for future work.

%Часть ННГУ - конец

\section{Mathematical model}

Let's consider the scheme of chemical transformations of the isomerization of the pentane-hexane fraction. The reaction consists of 48 reaction stages, ...

2-MP is 2-methyl pentane;

3-MP is 3-methyl pentane;

2,2-DMB is 2,2-dimethyl butane;

2,3-DMB is 2,3-dimethyl butane

CH* is cyclohexane

MCP is methyl cyclopentane

B is benzene.

3-MP is 3-methyl pentane;

2,2-DMB is 2,2-dimethyl butane;

2,3-DMB is 2,3-dimethyl butane

CH* is cyclohexane

\begin{table}[H] 
\caption{Transformations of the catalytic isomerization of the pentane-hexane fraction.\label{tab11}}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\textwidth}{CCCC}
\toprule
\textbf{No}	& \textbf{Reaction}	& \textbf{E, kJ/mol}     & \textbf{log($k_0$)} \\
\midrule
1	& $n$-C$_5$H$_{12}$ $\rightarrow$ $iso$-C$_5$H$_{12}$ & 149.320 & 18.492\\
2 & $iso$-C$_5$H$_{12}$ $\rightarrow$ $n$-C$_5$H$_{12}$ & 160.474 & 19.507 \\
3 & $n$-C$_6$H$_{14}$ $\rightarrow$ 2-MP & 102.833 & 8.127 \\
4 & 2-MP $\rightarrow$ $n$-C$_6$H$_{14}$ & 106.317 & 11.279 \\

\bottomrule
\end{tabularx}
\end{table}

The introduction should briefly place the study in a broad context and highlight why it is important. It should define the purpose of the work and its significance. The current state of the research field should be reviewed carefully and key publications cited. Please highlight controversial and diverging hypotheses when necessary. Finally, briefly mention the main aim of the work and highlight the principal conclusions. As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research. Citing a journal paper \cite{ref-journal}. Now citing a book reference \cite{ref-book1,ref-book2} or other reference types \cite{ref-unpublish,ref-communication,ref-proceeding}. Please use the command \citep{ref-thesis,ref-url} for the following MDPI journals, which use author--date citation: Administrative Sciences, Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, Journal of Intelligence, Journalism and Media, JRFM, Languages, Laws, Religions, Risks, Social Sciences, Literature.

%Часть ННГУ - начало
\section{Parallel Algorithm for Solving Global Optimization Problems }\label{Sec_GSA}

@@

As mentioned above, from a mathematical point of view, the inverse problem of chemical kinetics can be considered a Lipschitz global optimization problem. 
In general, the problem of the specified class can be formulated as follows:
\begin{gather}
 \phi^* = \phi(y^\ast)=\min{\left\{\phi(y):y\in D\right\}}, \label{problemN}\\
 D=\left\{y\in R^N: a_i\leq y_i \leq b_i, \;  1\leq i \leq N\right\} \label{D},
\end{gather}
where $a,b \in R^N$ are boundaries of the search domain, and the objective function $\phi(y)$ satisfies the Lipschitz condition
\begin{equation}\label{Lip}
\left|\phi(y_1)-\phi(y_2)\right|\leq L\left\|y_1-y_2\right\|,\; y_1,y_2 \in D.
\end{equation}

The function $\phi(y)$ is assumed to be multiextremal and is given in the form of some computational algorithm. The input of this algorithm is supplied with a parameter vector, and the output is the corresponding value of the function. Such functions are often referred to as ``black-box'' functions. It is also assumed that each calculation of the function value at the search domain point (hereinafter also referred to as trial) is a laborious operation and requires significant computational resources. As noted in the introduction, this statement of the problem fully corresponds to the inverse problem of chemical kinetics.

@@

Known examples of such methods are the non-uniform space covering method \cite{Evtushenko2013}, the simplicial partitions method \cite{PaulaviciusZilinskas2014}, the diagonal partitions method \cite{Sergeyev2017}. These techniques are based on adaptive partitioning of the search domain into a finite number of subdomains at each iteration.
Then the need for further subdivision for a more detailed study of these subdomains is estimated (using the accumulated information about the objective function). 
A fundamentally different approach to solving a multidimensional problem (\ref{problemN}) is its reduction to one or more one-dimensional problems followed by the use of efficient methods of one-dimensional optimization. Such reduction may be accomplished, for example, by a nested optimization scheme \cite{Grishagin2018} or by Peano-Hilbert curves \cite{Barkalov2018}. The latter approach is used in this work.

Using a continuous single-valued mapping (Peano-Hilbert curve) $y(x)$ of the real axis segment $[0,1]$ onto hyperinterval $D$ from (\ref{D}), a multidimensional problem (\ref{problemN}) can be reduced to a one-dimensional one
\[
\phi(y^\ast)=\phi(y(x^\ast))=\min{\left\{\phi(y(x)): x\in[0,1]\right\}}.
\]
It is known \cite{Strongin2000, Sergeyev2013} that the reduced one-dimensional function $\phi(y(x))$ will no longer be Lipschitzian, but will satisfy the H{\"o}lder condition
\[
\left|\phi(y(x_1))-\phi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N},
\]
where the H{\"o}lder constant $H$ is associated with the Lipschitz constant $L$ by the relationship $ H=2 L \sqrt{N+3}$.

@@

Thus, under this approach, the search trial at some point $x^k\in[0,1]$ will include the construction of the image $y^k=y(x^k)$ first, and only then the calculation of the value of the function $ z^k = \phi(y^k)$.
We will call the three values $\{x^k,y ^k,z^k\}$ the trial result.

%поправить название
\subsection{Asynchronous parallel algorithm used to determine the global minimum of a function}

The parallel algorithm used in this study is based on the information and statistical approach to the development of global optimization methods, the theoretical foundations of which are set out in \cite{Strongin2000}.

@@

Let us assume that at each iteration, the master-process calculates one point of the new trial and transfers it for testing to the worker-process. At the same time, a trial performed by a worker is a much more labor-intensive operation than selection of a new point by a master, which eliminates the downtime of worker-processes. 
In this case (unlike synchronous parallel algorithms), the total number of trials performed by each worker-process  will depend on the labor intensity of the specific trial and cannot be estimated in advance.

@@

The initial iteration is carried out according to a special rule. The master process (let us assume that this is the process No 0) initiates  $p$ parallel trials at $p$ different points of the search domain, two of which are boundary points, and the rest are arbitrary internal points (for example, uniformly distributed). Thus, the initial trials are carried out at the points $\{y(x^1), y(x^2), ...,y(x^p)\}$, where $x^1 = 0$, $x^p = 1$, $x^i\in(0,1), i=2,..., p-1$.

@@

Let one of the processes (without limiting the commonality, we can assume that this is the process No 1), at some point in time complete the trial at the point $y(x^{k+1})$.
The remaining processes are at the stage of conducting their trials, i.e. at these points multiple trials have already started, but have not yet been completed.

@@

Step 1. Renumber he point of the set 
\[
X_k = \left\{x^1, x^2,...,x^{k+p} \right\},
\]
which contains all the points at which the trials either have been carried out or are being carried out by the lower index in ascending order, i.e.
\[
0=x_1<x_2<...<x_{x+p}=1.
\]
Step 2. Calculate values
\[
M_1=\max \left\{ \frac{ \left|z_i - z_{i-1} \right|}{(x_i-x_{i-1})^{1/N}} : x_{i-1} \notin I_k, x_i \notin I_k, 2\leq i\leq k+p \right\},
\]
\[
M_2=\max \left\{ \frac{ \left|z_{i+1} - z_{i-1} \right|}{(x_{i+1}-x_{i-1})^{1/N}} : x_i \in I_k, 2\leq i < k+p \right\},
\]
\[
M=\max\{M_1,M_2\},
\]
where $z_i=\phi(y(x_i))$, if $x_i \notin I_k, \; 1\leq i \leq k+p$. The $z_i$ values at $x_i \in I_k$ are undefined because the trials at $x_i \in I_k$ are not yet complete. If the value $M$ is 0, assign $M=1$.

@@

Immediately after calculating the point $x^{k+p+1}$ of the next trial, the master-process adds it to the set $I_k$ and forwards it to the worker-process, which initiates the trial at this point.

The master process terminates the algorithm if either the condition $\Delta_{t}<\varepsilon$ ($t$ is number of the interval with the maximum characteristic from Step 4) or the condition $k>K_{max}$ ($k$ is the number of trials performed) is met.
The real number $\varepsilon>0$ and the integer $K_{max}>0$ are the algorithm parameters which correspond to the accuracy of the solution search and the maximum number of trials.

@@

\section{Numerical Experiments}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
%\supplementary{The following supporting information can be downloaded at:  \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.}

>>>>

%=================================================================
% Full title of the paper (Capitalized)
\Title{On Solving the Problem of Finding Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction Using Parallel Global Search Algorithm}
% Дежурное название: Использование ...алгоритма для определения кинетических параметров процесса каталитической изомеризации пентан-гексановой фракции
% Еще один вариант: О решении задачи определения Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction с помощью Parallel Global Search Algorithm
% На английском: On Solving the Problem of Finding Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction Using Parallel Global Search Algorithm

% MDPI internal command: Title for citation in the left column
\TitleCitation{On Solving the Problem of Finding Kinetic Parameters}

% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0000-0001-5273-2471} % Add \orcidA{} behind the author's name
\newcommand{\orcidauthorB}{0000-0001-6776-0096}
\newcommand{\orcidauthorC}{0000-0002-8736-0652} % Add \orcidB{} behind the author's name
\newcommand{\orcidauthorD}{0000-0001-6073-9585} 
\newcommand{\orcidauthorE}{0000-0003-4219-4870}

% Authors, for the paper (add full first names)
\Author{Konstantin Barkalov $^{1}$\orcidA{}, Irek Gubaydullin $^{2,3}$, Evgeny Kozinov$^{1}$\orcidB{}, Ilya Lebedev $^{1}$\orcidC{}, Roza Faskhutdinova\orcidD{} $^{2,3}$, Azamat Faskhutdinov$^{2}$ and Leniza Enikeeva $^{3,4,}$*\orcidE{}}

@@

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{Konstantin Barkalov, Irek Gubaydullin, Evgeny Kozinov, Ilya Lebedev, Roza Faskhutdinova, Azamat Faskhutdinov, Leniza Enikeeva}

% MDPI internal command: Authors, for citation in the left column
\AuthorCitation{Barkalov, B.; Gubaydullin, I.; Kozinov, E.; Lebedev, I.; Faskhutdinova, R.; Faskhutdinov, A.; Enikeeva, L.}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{The article is devoted to the problem of developing a kinetic model of a complex chemical reaction using a parallel optimization method. The design of the kinetic model consists in finding the kinetic parameters of the reaction, which cannot be calculated analytically, and since the chemical reaction involves many stages, the optimization problem is multi-extreme. As a chemical reaction, the process of catalytic isomerization of the pentane-hexane fraction is considered, which is now important due to the switch of the oil refining industry to the production of gasoline corresponding to the Euro-5 standard. To solve the optimization problem, the authors have developed parallel global search algorithm and a program based on Lipschitz optimization. The kinetic parameters found made it possible to develop a mathematical model of the process, which gives good agreement with industrial data.}

% Keywords
\keyword{kinetic model; inverse problems of chemical kinetics; mathematical modeling; catalytic isomerisation of pentane-hexane fraction; Lipschitz optimization; black-box functions; parallel computing}

@@

Due to the transition of the domestic oil refining industry to the production of gasoline conforming to the Euro-5 standard, an urgent task is to reduce the content of aromatic hydrocarbons and, in particular, benzene in motor fuel while maintaining the octane value. In connection with the transition of the oil refining industry to the production of motor gasoline conforming to Euro-5 and Euro-6 standards, the reduction of aromatic hydrocarbons in motor fuel and, mainly, benzene becomes an urgent task. At the same time, removing aromatic hydrocarbons, which are high-octane components, it is necessary to preserve the octane number of gasoline. It is known that hydrocarbons of branched structure have a higher octane number than hydrocarbons of linear structure. For example, in $n$-pentane, the octane number according to the research method is 62 points, and in $iso$-pentane it is 93 points. On an industrial scale, catalytic isomerization units of light paraffins have been introduced to produce hydrocarbons of an isomeric structure. The catalytic isomerization of light paraffins, in turn, makes it possible to obtain a high-octane component of automobile gasoline with a minimum content of aromatic hydrocarbons. The main advantage of this process is the production of high-octane gasoline, while aromatic hydrocarbons are practically not formed. Straight-run gasoline fractions with boiling limits of 62-70 $^0$C containing mainly $n$-pentane and $n$-hexane, pentane-hexane fractions from gas condensate or from gas fractionation plants, as well as catalytic reforming raffinates are used as feedstock for catalytic isomerization. Modern bifunctional catalysts are used to carry out the process. In order to prevent coke deposition, the catalytic isomerization process is carried out in a hydrogen-containing gas environment under pressure.

Computer information technologies are widely used in the modeling of production processes and today it is difficult to imagine research activities without the use of modern means of constructing and using models. Moreover, special attention is currently being paid to the development of a detailed kinetic model of complex industrial processes. Based on these models, it is possible to significantly increase the efficiency of the process, while minimally spending energy and material resources. The main advantage of such methods is the absence of additional costs for the construction of pilot plants. The developed model should describe the process within a wide range of operating conditions, have high flexibility, allowing for changes, calculations of the composition of the reaction mixture, etc.

In the article \cite{enik2021} we have already reviewed this process, but in order to simplify the model, the scheme of chemical transformations, in particular hydrocracking reactions, have been revised. Thus, the most probable hydrocracking reactions based on the carbonium-ion mechanism were formed. The new transformation scheme implies the recalculation of the kinetic parameters of the reaction, which form a mathematical model of the process.

In the article \cite{enik2021}, a genetic algorithm was used to calculate kinetic parameters, and the calculation time of kinetic parameters was about 10 hours. Therefore, in order to develop a mathematical model of a new transformation scheme, the question of choosing a more effective optimization method arises.

@@

Computational algorithms for solving such multiextremal problems (global optimization methods) differ vastly from local optimization techniques since they require the study of the entire search domain (see, for example, \cite{PaulaviciusZilinskas2014,Sergeyev2017}). These methods can be divided into two classes: metaheuristic and deterministic. Metaheuristic algorithms are usually based on simulations of processes occurring in the wild.
Examples of metaheuristic algorithms are simulated annealing, evolutionary computing, etc. (see, for example, \cite{Battiti2009,Eiben2015}).
These algorithms are included in many software libraries for solving global optimization problems (for example, MATLAB Global Optimization Toolbox, Scipy.Optimize, etc.).
Due to the operation speed and low requirements for computing resources (in particular, memory), metaheuristic algorithms have become widespread for solving practical problems. However, the solution to the problem found by such an algorithm is, in general, local and can be located far from the global optimum \cite{Kvasov2018}.

@@

This paper reflects the results of applying parallel methods of Lipschitz optimization developed in \cite{Barkalov2016,Strongin2018,Sysoyev2017}, to the solution of inverse problems of chemical kinetics.  
The main part of the paper is structured as follows.

Section \ref{sec_matmodel} provides a description of the mathematical model of the chemical reaction under study. Section \ref{sec_GSA} discusses the standard formulation of the Lipschitz global optimization problem and the algorithm for solving it. Section \ref{sec_numexper} outlines the issues related to the technical aspects of the implementation of the parallel algorithm. The physico-chemical interpretation of the results of numerical calculations using the developed parallel algorithm is presented in the Section \ref{sec_discussion}. 
%The last section contains the conclusion and plans for future work.

\section{Mathematical model}\label{sec_matmodel}

Let's consider the mechanism of chemical transformations of the isomerization of the pentane-hexane fraction. The reaction consists of 48 reaction stages, which are presented in the table \ref{tab_chemistry}. The following abbreviations were adopted: 2-MP is 2-methylpentane; 3-MP is 3-methylpentane; 2,2-DMB is 2,2-dimethylbutane; 2,3-DMB is 2,3-dimethylbutane; CH* is cyclohexane; MCP is methyl cyclopentane; B is benzene; CP is cyclopentane.

\begin{table}[H] 
\caption{Chemical transformations of the catalytic isomerization of the pentane-hexane fraction.\label{tab_chemistry}}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\textwidth}{CC}
\toprule
 \textbf{Reaction stages} & \textbf{Reaction stages} \\
\midrule
1) $n$-C$_5$H$_{12}$ $\rightarrow$ $iso$-C$_5$H$_{12}$ & 25) $n$-C$_6$H$_{14}$ $\rightarrow$ CH* + H$_2$\\
2) $iso$-C$_5$H$_{12}$ $\rightarrow$ $n$-C$_5$H$_{12}$ & 26) CH* + H$_2$ $\rightarrow$ $n$-C$_6$H$_{14}$ \\
3) $n$-C$_6$H$_{14}$ $\rightarrow$ 2-MP & 27) 2-MP $\rightarrow$ MCP+ H$_2$ \\
4) 2-MP $\rightarrow$ $n$-C$_6$H$_{14}$ & 28) MCP + H$_2$ $\rightarrow$ 2-MP  \\
5) $n$-C$_6$H$_{14}$ $\rightarrow$ 3-MP & 29) 3-MP $\rightarrow$ MCP + H$_2$ \\
6) 3-MP $\rightarrow$ $n$-C$_6$H$_{14}$  & 30) MCP + H$_2$ $\rightarrow$ 3-MP \\
7)  2-MP $\rightarrow$ 3-MP & 31) 2,2-DMB $\rightarrow$ MCP + H$_2$ \\
8) 3-MP $\rightarrow$ 2-MP  & 32) MCP + H$_2$ $\rightarrow$ 2,2-DMB \\
9) 2-MP $\rightarrow$ 2,2-DMB  & 33) 2,3-DMB $\rightarrow$ MCP + H$_2$ \\
10) 2,2-DMB $\rightarrow$ 2-MP  & 34) MCP + H$_2$ $\rightarrow$ 2,3-DMB \\
11) 2-MP $\rightarrow$ 2,3-DMB  & 35) CP + H$_2$ $\rightarrow$ $n$-C$_5$H$_{12}$ \\
12) 2,3-DMB $\rightarrow$ 2-MP  & 36) $n$-C$_5$H$_{12}$ $\rightarrow$ CP + H$_2$ \\
13) 3-MP $\rightarrow$ 2,2-DMB  & 37) $n$-C$_4$H$_{10}$ $\rightarrow$ $iso$-C$_4$H$_{10}$ \\
14) 2,2-DMB $\rightarrow$ 3-MP  & 38) $iso$-C$_4$H$_{10}$ $\rightarrow$ $n$-C$_4$H$_{10}$ \\
15) 3-MP $\rightarrow$ 2,3-DMB  & 39) $n$-C$_5$H$_{12}$ + H$_2$ $\rightarrow$ C$_3$H$_8$ + C$_2$H$_6$ \\
16) 2,3-DMB $\rightarrow$ 3-MP  & 40) $n$-C$_5$H$_{12}$ + H$_2$ $\rightarrow$ $n$-C$_4$H$_{10}$ + CH$_4$ \\
17) 2,2-DMB $\rightarrow$ 2,3-DMB  & 41) $iso$-C$_5$H$_{12}$ + H$_2$ $\rightarrow$ $iso$-C$_4$H$_{10}$ + CH$_4$ \\
18) 2,3-DMB $\rightarrow$ 2,2-DMB & 42) $n$-C$_6$H$_{14}$ + H$_2$ $\rightarrow$ 2C$_3$H$_8$ \\
19) B + 3H$_2$ $\rightarrow$ CH*  & 43) $n$-C$_6$H$_{14}$ + H$_2$ $\rightarrow$ $n$-C$_5$H$_{12}$ + CH$_4$ \\
20) CH* $\rightarrow$ B + 3H$_2$ & 44) $n$-C$_6$H$_{14}$ + H$_2$ $\rightarrow$ $n$-C$_4$H$_{10}$ + C$_2$H$_6$ \\
21) B + 3H$_2$ $\rightarrow$ MCP  & 45) 2-MP + H$_2$ $\rightarrow$ $iso$-C$4$H$_{10}$ + C$_2$H$_6$ \\
22) MCP $\rightarrow$ B + 3H$_2$  & 46) 3-MP + H$_2$ $\rightarrow$ $iso$-C$_5$H$_{12}$ + CH$_4$ \\
23) CH* $\rightarrow$ MCP & 47) 2,2-DMB + H$_2$ $\rightarrow$ $iso$-C$_5$H$_{12}$ + CH$_4$ \\
24) MCP $\rightarrow$ CH*  & 48) 2,3-DMB + H$_2$ $\rightarrow$ $iso$-C$_5$H$_{12}$ + CH$_4$ \\
\bottomrule
\end{tabularx}
\end{table}

Mathematical model of the chemical process is a non-linear system of differential equations \cite{Koledina2016KineticsAM,enik2021,logan}:
\begin{linenomath}
\begin{equation}\label{eq_concentrations}
\dfrac{dx_i}{d\tau} = \sum_{j=1}^J \nu_{ij} w_j, i = 1,...,I,
\end{equation}
\begin{equation}\label{eq_rate}
w_{j} = k_j \prod_{i=1}^I \left( \dfrac{x_i}{F} \right),
\end{equation}
\begin{equation}\label{eq_constant}
k_j = k^0_j \exp \left( -\dfrac{E_j}{RT} \right)
\end{equation}
\end{linenomath}
with initial values: at $\tau = 0$, $x_i(0) = x_i^0$, where $\nu_{ij}$ are stoichiometric coefficients of stages of chemical reaction; $J$ is number of stages ($J = 48$), $x_i$ is molar flow rate of the $i$-th component of reaction, kmol/h; $I$ is the number of components ($I = 17$); $w_j$ is the rate of the $j$-th stage of reaction, kmol/(h$\cdot$kg cat.), $E_j$ is activation energy of reaction stage, J/mol; $R$ is the universal gas constant ($R$=8,31 J/(mol$\cdot$K), $T$ is temperature, K; $k_j$ are pre-exponentional factors, kmol/(h$\cdot$kg cat.); $\tau$ is conditioned contact time, (kg cat.); $F$ is the total molar flow rate, kmol/h.

The mathematical model involves such parameters as the raw material flow rate $F$ in the reactor and temperature $T$, but these two parameters also depend on $\tau$, therefore it is necessary to write equations on them, because technological process is non-isothermal:
\begin{linenomath}
\begin{equation}\label{eq_temper}
\dfrac{dT}{d\tau} = \dfrac{\sum_{i=1}^I\left( \dfrac{dx_i}{d\tau} \cdot \Delta H^i_{f,T} \right) }{\sum_i^J \left(x_i C^T_{p_i} \right) },
\end{equation}\label{eq_flow}
\begin{equation}
\dfrac{dF}{d\tau} = \sum_{i=1}^I  \dfrac{x_i}{d\tau}
\end{equation}
\end{linenomath}
with initial condition: at $\tau = 0$, $T(0) = T^0$, where $T$ is temperature, K, $\Delta H^i_{f,T}$ is enthalpy of the $i$-th component formation at temperature $T$, J/mol, $C^T_{p_i}$ is specific heat capacity of the $i$-th component at temperature $T$, (J/mol$\cdot$K), $F$ is mole discharge of the flow, kmol/h.

Thus, when solving the inverse kinetic problem, the direct problem is repeatedly solved, which includes the numerical integration of equations \eqref{eq_concentrations}, \eqref{eq_temper} and \eqref{eq_flow}. After finding the numerical values of the molar flow rate $x_i$, temperature $T$ and total molar flow rate $F$, these values are compared with the corresponding industrial data obtained during the three days of operation of the catalytic isomerization unit \cite{enik2021}, which is a minimization task:

\begin{linenomath}
\begin{equation}
FF = \dfrac{\sum_{i=1}^I \left| \dfrac{y_i^{exp} - y_i^{calc}}{y_i^{exp}} \right|}{I} \rightarrow \min,
\end{equation}
\end{linenomath}

where I is the number of observed parameters (I=19, of which 17 parameters are $x_i$, one parameter is $T$ and one parameter is $F$),  $y_{i}^{exp}$ are industrual data, $y_{i}^{calc}$ are numerical values of concentrations of reaction substances obtained by solving a direct kinetic problem. As a result of solving the inverse problem, the kinetic parameters $k^0_j$ and $E_j$ included in the Arrhenius equation \eqref{eq_rate} will be obtained. Since the number of reaction stages is 48, the kinetic model will contain 96 kinetic parameters (activation energy $E_j$ and pre-exponential multiplier $k^0_j$ for each stage), however, the authors decided to fix some of the parameters as already known for chemical reasons, so the number of optimized parameters was reduced to 24.

\section{Parallel Algorithm for Solving Global Optimization Problems }\label{sec_GSA}

@@

As stated above, from a mathematical point of view, the inverse problem of chemical kinetics can be considered a Lipschitz global optimization problem. 
In general, the problem of the specified class can be formulated as follows:
\begin{gather}
 \phi^* = \phi(y^\ast)=\min{\left\{\phi(y):y\in D\right\}}, \label{problemN}\\
 D=\left\{y\in R^N: a_i\leq y_i \leq b_i, \;  1\leq i \leq N\right\} \label{D},
\end{gather}
where $a,b \in R^N$ are boundaries of the search domain, and the objective function $\phi(y)$ satisfies the Lipschitz condition
\begin{equation}\label{Lip}
\left|\phi(y_1)-\phi(y_2)\right|\leq L\left\|y_1-y_2\right\|,\; y_1,y_2 \in D.
\end{equation}

The criterion $\phi(y)$ is assumed to be multiextremal and is given in the form of some computational algorithm. The input of this algorithm is supplied with a parameter vector, and the output is the corresponding function value. Such functions are often referred to as ``black-box'' functions. It is also assumed that each calculation of the function value at the search domain point (hereinafter also referred to as trial) is a laborious operation and requires significant computational resources. As noted in the introduction, this statement of the problem fully corresponds to the inverse problem of chemical kinetics.

@@

Known examples of such methods are DIvide RECTangles (DIRECT) algorithm \cite{Jones1993}, the simplicial partitions algorithm \cite{PaulaviciusZilinskas2014}, the diagonal partitions algorithm \cite{Sergeyev2017}. These techniques are based on adaptive partitioning of the search domain into a finite number of subdomains at each iteration.
Then the need for further subdivision for a more detailed study of these subdomains is estimated (using the accumulated information about the objective function). 
A fundamentally different approach to solving a multidimensional problem (\ref{problemN}) is its reduction to one or more one-dimensional problems followed by the use of efficient methods of one-dimensional optimization. Such reduction may be accomplished, for example, by a recursive optimization scheme \cite{Grishagin2018} or by Peano curves \cite{Barkalov2018}. The latter approach is used in this work.

Using a continuous single-valued correspondence (Peano curve) $y(x)$ of the real axis segment $[0,1]$ onto hyperinterval $D$ from (\ref{D}), a multidimensional problem (\ref{problemN}) can be reduced to a one-dimensional one
\[
\phi(y^\ast)=\phi(y(x^\ast))=\min{\left\{\phi(y(x)): x\in[0,1]\right\}}.
\]
It is known \cite{Strongin2000, Sergeyev2013} that the reduced one-dimensional function $\phi(y(x))$ will no longer be Lipschitzian, but will satisfy the H{\"o}lder condition
\[
\left|\phi(y(x_1))-\phi(y(x_2))\right|\leq H\left|x_1-x_2\right|^{1/N},
\]
where the H{\"o}lder constant $H$ is associated with the Lipschitz constant $L$ by the formula $ H=2 L \sqrt{N+3}$.

@@

Thus, under this approach, the search trial at some point $x^k\in[0,1]$ will consist of constructing the image $y^k=y(x^k)$ first, and only then of calculating the function value $ z^k = \phi(y^k)$.
We will call the three values $\{x^k,y ^k,z^k\}$ the trial result.

\subsection{Asynchronous parallel algorithm used to determine the global minimum of a function}

The parallel algorithm used in this study is based on the information-statistical approach to the development of global optimization methods, the theoretical foundations of which are set out in \cite{Strongin2000}.

@@

Let us assume that at each iteration, the master calculates one new trial  point and transfers it for processing to the worker. At the same time, a trial performed by a worker is a much more labor-intensive operation than selection of a new point by a master, which eliminates the worker downtime. 
In this case, the total number of trials performed by each worker  will depend on the labor intensity of the specific trial and cannot be estimated in advance.

@@

The initial iteration is carried out according to a special rule. The master process (let us assume that this is the process No 0) initiates  $p$ parallel trials at $p$ different points of the search region, two of which are boundary points, and the rest are arbitrary internal points (for example, uniformly distributed). Thus, the initial trials are carried out at the points $\{y(x^1), y(x^2), ...,y(x^p)\}$, where $x^1 = 0$, $x^p = 1$, $x^i\in(0,1), i=2,..., p-1$.

@@

Let one of the processes (without limiting the commonality, we can assume that this is the process No 1), at some point in time complete the trial at the point $y(x^{k+1})$.
The remaining processes are at the stage of conducting their trials, i.e. at these points multiple trials have already initiated, but have not yet been finalized.

@@

Step 1. Renumber he point of the set 
\[
X_k = \left\{x^1, x^2,...,x^{k+p} \right\},
\]
which contains all the points at which the trials either have been carried out or are being carried out by the lower index in ascending order, i.e.
\[
0=x_1<x_2<...<x_{x+p}=1.
\]
Step 2. Calculate values
\[
M_1=\max \left\{ \frac{ \left|z_i - z_{i-1} \right|}{(x_i-x_{i-1})^{1/N}} : x_{i-1} \notin I_k, x_i \notin I_k, 2\leq i\leq k+p \right\},
\]
\[
M_2=\max \left\{ \frac{ \left|z_{i+1} - z_{i-1} \right|}{(x_{i+1}-x_{i-1})^{1/N}} : x_i \in I_k, 2\leq i < k+p \right\},
\]
\[
M=\max\{M_1,M_2\},
\]
where $z_i=\phi(y(x_i))$, if $x_i \notin I_k, \; 1\leq i \leq k+p$.  If the value $M$ is 0, assign $M=1$. The values $z_i=\phi(y(x_i))$ at $x_i \in I_k$ are undefined because the trials at $x_i \in I_k$ are not finished yet.

@@

Immediately after calculating the point $x^{k+p+1}$ of the next trial, the master appends it to the set $I_k$ and forwards it to the worker, which starts the trial at this point.

The master stops the algorithm if either the condition $\Delta_{t}<\varepsilon$ (where $t$ is number of the interval with the maximum characteristic from Step 4) or the condition $k>K_{max}$ (where $k$ is the number of trials performed) is met.
The real number $\varepsilon>0$ and the integer $K_{max} \geq 1$ are the algorithm parameters which correspond to the accuracy of the solution search and the maximum number of trials.

@@

\section{Numerical Experiments}\label{sec_numexper}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
